What to simulate? Inferring the right direction for mental rotation

### 2014-02-05 man/cogsci-2014/cogsci-2014.tex
Update discussion

\input{../../results/D/overall_response_time.tex}
\input{../../results/D/overall_accuracy.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy_corrs.tex}
\input{../../results/D/human_corrs.tex}

@@

The average RT across all stimuli was \ExpTime{} milliseconds.  The
minimum angle of rotation was significantly monotonically correlated
with average per-stimulus RTs, both for ``flipped'' and ``same'' pairs
\TODO{}. While this replicates the general result of previous
experiments \cite<e.g.,>{Shepard1971,Cooper:1975wp}, our results are
not as perfectly linear. However, \citeA{Gardony:2013gn} found a
constant effect for ``flipped'' pairs; an average of their results and
the classic results would produce something similar to ours.

The average accuracy across all stimuli was \ExpAccuracy{}, though
there were \ExpNumChance{} stimuli (out of 720) for which people were
not above chance.  The minimum angle was also correlated with
participants' average per-stimulus accuracy, though significantly more
so for ``same'' pairs (\TODO{}). This is the same result found both by
\citeA{Cooper:1975wp} and \citeA{Gardony:2013gn}.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{../../figures/D/response_time_stimulus.pdf}
    \caption{\textbf{Typical RT curves for a single shape.}  These
      plots correspond to the shape shown in
      Fig.~\ref{fig:model-traces}. Left: human curves are either
      linear (as with the ``same'' pairs), or linear and then flat (as
      with the ``flipped'' pairs). Middle: the \Th{} model tends to
      have highly inconsitent within-shape RTs. Right: the \Bq{} model
      is roughly linear.}
    \label{fig:response-time-stimulus}
  \end{center}
\end{figure}

@@

Because the \Bq{} model relies on Eq.~\ref{eq:lh-ratio} for its
stopping criteria (as opposed to just finding a maximum), the prior
$p(h)$ has a fairly significant effect, as shown in
Fig.~\ref{fig:response-time-accuracy}. With just a small bias of
$p(h=0)=0.55$, there is a clear separation in RTs for ``same'' versus
``flipped'' stimuli; this separation is similar to the trend also
observed in human RTs. While the prior also had an effect on accuracy,
this effect did not reflect the trends in human behavior.

@@

We set out to answer the question of how people decide \textit{what}
to simulate when using mental imagery. Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment. While it is clear that none of the models
fully captured human behavior on the mental rotation task, it is worth
noting where they shed light on aspects that have previously gone
unexplained.  For example, the \Bq{} model's adaptive stopping rule is
sensitive to the prior, and thus provides a possible explanation for
why people are slower to respond on ``flipped'' stimulus pairs.

@@

How could the models we evaluated be improved? First, all of the
models use holistic rotations, however there is evidence that people
might compare individual features of shapes rather than the entire
shape at once \cite{Just1976,Yuille:1982tx}. Second, there are other
possible ``active sampling'' approaches that could combine the best
features from the \Th{} and \Bq{} models, such as maintaining a
distribution over the location and value of the global maximum, rather
than over the integral. Third, studying rotations with 3D objects of
varying levels of complexity may be more informative than the
extremely simple 2D objects used in this experiment.

As a parting comment, we note that the question of how people decide
\textit{which} rotations to perform is more difficult to answer than
previously assumed. Although we examined four well-motivated models,
none came close to explaining the full range of human behavior on this
task. We suggest that such metacognitive questions are fertile areas
for future research.

>>>>

\input{../../results/D/overall_response_time.tex}
\input{../../results/D/overall_accuracy.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy_corrs.tex}
\input{../../results/D/human_corrs.tex}
\input{../../results/D/theta_time_corrs.tex}
\input{../../results/D/theta_accuracy_corrs.tex}

@@

The average RT across all stimuli was \ExpTime{} milliseconds.  The
minimum angle of rotation was significantly monotonically correlated
with average per-stimulus RTs, both for ``flipped''
(\ExpThetaTimeCorrFlipped{}) and ``same'' pairs
(\ExpThetaTimeCorrSame{}). While this replicates the general result of
previous experiments \cite<e.g.,>{Shepard1971,Cooper:1975wp}, our
results are not as perfectly linear. However, \citeA{Gardony:2013gn}
found a constant effect for ``flipped'' pairs; an average of their
results and the classic results would produce something similar to
ours.

The average accuracy across all stimuli was \ExpAccuracy{}, though
there were \ExpNumChance{} stimuli (out of 720) for which people were
not above chance.  The minimum angle was also correlated with
participants' average per-stimulus accuracy, though significantly more
so for ``same'' pairs (\ExpThetaAccuracyCorrSame{}) than ``flipped''
pairs (\ExpThetaAccuracyCorrFlipped{}). This is the same result found
both by \citeA{Cooper:1975wp} and \citeA{Gardony:2013gn}.

@@

\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{../../figures/D/response_time_stimulus.pdf}
    \caption{\textbf{Typical RT curves for a single shape.}  These
      plots correspond to the shape shown in
      Fig.~\ref{fig:model-traces}. Left: human curves are either
      linear (as with the ``same'' pairs), or linear and then flat (as
      with the ``flipped'' pairs). Middle: the \Th{} model does not
      have a reliably monotonic relationship with rotation. Right: the
      \Bq{} model is roughly linear.}
    \label{fig:response-time-stimulus}
  \end{center}
\end{figure}

@@

Because the \Bq{} model relies on Eq.~\ref{eq:lh-ratio} for its
stopping criteria (as opposed to just finding a maximum), the prior
$p(h)$ had a fairly significant effect, as shown in
Fig.~\ref{fig:response-time-accuracy}. With just a small bias of
$p(h=0)=0.55$, there was a clear separation in RTs for ``same'' versus
``flipped'' stimuli; this separation is similar to the trend also
observed in human RTs. While the prior also had an effect on accuracy,
this effect did not reflect the trends in human behavior.

@@

We set out to answer the question of how people decide \textit{what}
to simulate when using mental imagery. Focusing on the specific case
of determining the direction and extent of mental rotation, we
formalized four models and compared their performance with the results
of a behavioral experiment.

The \Oc{} model was the best predictor of human RTs, but is an
unsatisfying explanation of human behavior because it relies on
\textit{a prior} knowledge that people are unlikely to have about
which points on the two shapes correspond. Similarly, while the \Th{}
was next best in terms of correlation, it makes the unsatisfying
assumption that the global maximum is known. Finally, neither of these
models can explain the difference in people's behavior on ``same'' and
``flipped'' stimuli.  The \Bq{} model was nearly as good as the \Th{}
model; in contrast to the other models, the \Bq{} model makes no
assumptions about people's \textit{a prior} knowledge. Furthermore,
the \Bq{} model's adaptive stopping rule is sensitive to the prior,
and thus provides a possible explanation for why people are slower to
respond on ``flipped'' stimulus pairs.

@@

Thus, we suggest that the \Bq{} model offers the most promising
explanation of people's behavior on the mental rotation task to
date. It is the only model out of those surveyed which did not make
assumptions about people's \textit{a priori} knowledge, which was a
reasonable predictor of people's behavior, and which provided an
explanation for the longer response times on ``flipped'' stimuli.
Clearly, though, the \Bq{} model is not a perfect account, and there
are several immediately obvious ways in which it could be
improved. For example, while we used holistic rotations in this paper,
there is evidence that people compare individual features of shapes
rather than the entire shape at once
\cite{Just1976,Yuille:1982tx}. Second, there are other possible active
sampling approaches that could combine the best features from the
\Th{} and \Bq{} models, such as maintaining a distribution over the
location and value of the global maximum, rather than over the
integral.

To conclude, we hypothesize that people decide \textit{what} to
simulate through an active sampling strategy similar to that used by
the \Bq{} model. While more research is required to confirm or
disprove this claim, the work presented in this paper provides the
foundation necessary to do so.
